---
title: "DEG and Gene Ontology enrichment"
author: "Goran Royer"
date: "2026-09-25"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r Libraries}

packages <- c("Seurat","dplyr","tibble","AnnotationHub","AnnotationDbi",
              "clusterProfiler","ggplot2","enrichplot","dplyr",
              "openxlsx", "readxl")
invisible(lapply(packages, library, character.only = TRUE))
```


```{r Function and Dataset, echo=FALSE}

source(".../Script_utility/tau_index.R")

atlas <- readRDS(".../ATLAS_ASTMEX.RDS")
se <- subset(atlas, col_tissue == "tissue") # replace col_tissue by the tissue column of your interest

```


## Differentially Expressed Gene (DEG) analysis

In the Seurat package, two functions are dedicated to the analysis of the differential gene expression between two conditions, clusters, ...
'FindAllMarkers()' allows to finds markers for a specific identity compared to the rest of your dataset.
'FindMarkers()' is quite similar to the previous function except for one particular feature. You can decide which groups you want to compare. 
You will get two Excel files. A first excel file for the raw data and a second one containing the significant, filtered and specific DEGs

```{r Differential expressed genes analysis}

# 1. Extract the different timepoints
stages <- unique(se$age)

# 2. Identify the genes whose expression differs between the two conditions
df_deg <- list()
for (i in stages) {
  print(i)
  sub <- subset(se,age %in% i)
  Idents(sub) <- "condition"
  markers <- FindMarkers(
    object = sub,
    ident.1 = "surface", 
    ident.2 = "cave",
    logfc.threshold = 0.5)
  
  markers <- rownames_to_column(markers, var = "gene")
  df_deg[[i]] <- markers
  rm(markers, sub)
}

# 3. Raw data
wb <- createWorkbook()
for (j in names(df_deg)) {
  addWorksheet(wb, sheetName = j)
  writeData(wb, sheet = j, x = df_deg[[j]], colNames = TRUE, rowNames = FALSE)
}
## Obtain an excel containing a sheet for markers of each stage
saveWorkbook(wb, "results_path/DEG_by_stage_raw.xlsx", overwrite = TRUE)


# 4. Keep the genes whose the p-value adjusted is less than 0.05)
df_deg <- setNames(lapply(stages, function(x)
  filter(df_deg[[x]], p_val_adj <= 0.05)
),stages)


# 5. DEG filtered
df_deg.filt <- setNames(lapply(stages, function(x)
  filter(df_deg[[x]], (pct.1 >= 0.1 | pct.2 >= 0.1))
), paste0(stages, "_filt"))


# 6. DEG_spe + filtered ----
## Only used on a tissue
spe_tissue <- setNames(lapply(stages, function(x) {
  subset(atlas, age %in% x) |>
    calculate_tau_seurat(group_by = col_tissue, min_expr = 0.1) |>
    get_specific_genes(tau_threshold = 0.8, top_n = NULL)
}), stages)
## You will get a gene list for each tissue in your column

list_genes_spe <- sapply(stages, function(x) 
  intersect(spe_germlayers[[x]]$tissue$gene, df_deg[[x]]$gene),
  USE.NAMES = TRUE)

df_spe <- setNames(mapply(function(df, genes)
  df[df$gene %in% genes, ], 
  df_deg.filt, list_genes_spe, SIMPLIFY = FALSE), 
  paste0(stages,"_spe")) ## We retrieve the DEG table containing only the specific genes

df_final <- append(df_deg, c(df_deg.filt, df_spe))

wb <- createWorkbook()
for (j in names(df_final)) {
  addWorksheet(wb, sheetName = j)
  writeData(wb, sheet = j, x = df_final[[j]], colNames = TRUE, rowNames = FALSE)
}
saveWorkbook(wb, "results_path/DEG_by_stage.xlsx", overwrite = TRUE)
## sheet ending "_filt" correspond to DEGs filtered, and those ending in ‘_spe’ to specific DEGs
```


## Gene Ontology enrichment

To start we need to get the gff file on the NCBI ('https://www.ncbi.nlm.nih.gov/datasets/taxonomy/7994/'). 
It contains all the annotation features. 

```{r Mapping dataframe}
# Read and process GFF file
gff <- read.table(".../ncbi_dataset/data/GCF_023375975.1/genomic.gff", 
                  sep="\t",
                  comment.char = "#",
                  quote = "",
                  stringsAsFactors = FALSE
)
colnames(gff) <- c("seqid","source","type","start","end","score","strand","phase","attributes")

# Keep only genes and pseudogenes
gff_genes <- gff %>%
  filter(type == "gene" | type == "pseudogene")

# Extract GeneID and gene names
geneID <- ifelse(
  grepl("GeneID:", gff_genes$attributes),
  sub(".*GeneID:([^;,]+).*", "\\1", gff_genes$attributes),
  NA
)

geneName <- ifelse(
  grepl("gene=", gff_genes$attributes),
  sub(".*gene=([^;]+).*", "\\1", gff_genes$attributes),
  NA
)

# Create mapping dataframe
gff_ID <- data.frame(geneName = geneName, GeneID = geneID, stringsAsFactors = FALSE)
rm(geneName, geneID, gff_genes)
```

```{r universe}

universe_entrez <- unique(gff_ID$GeneID)
```

```{r annotation database}
hub <- AnnotationHub()
org.ast <- hub[["AH119686"]]

#check
gene_ids <- keys(org.ast, keytype="ENTREZID")

# Verification
length(intersect(universe_entrez, gene_ids))
length(universe_entrez)
length(gene_ids)

## Even though the size of `gene_ids` is larger, the most important thing is that the number of genes shared between `gene_ids` and `universe_entrez` is equal to the size of `universe_entrez`
```


```{r}
enrichment <- function(deg_path) {
  sheets <- getSheetNames(deg_path)
  sheets_all <- sheets[!grepl(pattern = "_filt|_spe", x = sheets)]
  sheets_filt <- sheets[grepl(pattern = "_filt", x = sheets)]
  sheets_spe <- sheets[grepl(pattern = "_spe", x = sheets)]
  
  # Initialiser uniquement les listes nécessaires
  ego_results_all <- list()
  ego_simpl_all <- list()
  ego_results_filt <- list()
  ego_simpl_filt <- list()
  
  # Traitement des feuilles "all"
  for(stage in sheets_all) {
    DEG_data <- read.xlsx(deg_path, sheet = stage, rowNames = FALSE)
    if (nrow(DEG_data) <= 1) {
      message(paste("Feuille", stage, "est vide. Ignorée."))
      next
    }
    DEG_data$gene <- as.character(DEG_data$gene)
    deg_map <- DEG_data %>% left_join(gff_ID, by = c("gene" = "geneName"))
    
    deg_entrez <- na.omit(deg_map$GeneID)
    
    ego <- enrichGO(gene = deg_entrez, 
                    universe = universe_entrez, 
                    OrgDb = org.ast, 
                    keyType = "ENTREZID", 
                    ont = "BP",
                    pvalueCutoff = 0.05, 
                    pAdjustMethod = "BH",
                    qvalueCutoff = 0.2, 
                    readable = TRUE)
    ego_results_all[[stage]] <- ego
    if (!is.null(ego) && nrow(ego@result) > 0) {
      message(paste("Nombre de GO terms enrichis pour", stage, ":", nrow(ego@result)))
    } else {
      message(paste("Aucun GO term significatif trouvé pour", stage))
    }
    ego_simpl_all[[stage]] <- simplify(ego)
  }
  
  # Traitement des feuilles "_filt"
  for(stage in sheets_filt) {
    DEG_data <- read.xlsx(deg_path, sheet = stage, rowNames = FALSE)
    if (nrow(DEG_data) <= 1) {
      message(paste("Feuille", stage, "est vide. Ignorée."))
      next
    }
    DEG_data$gene <- as.character(DEG_data$gene)
    deg_map <- DEG_data %>% left_join(gff_ID, by = c("gene" = "geneName"))
    
    deg_entrez <- na.omit(deg_map$GeneID)
    
    ego <- enrichGO(gene = deg_entrez, 
                    universe = universe_entrez, 
                    OrgDb = org.ast, 
                    keyType = "ENTREZID", 
                    ont = "BP", 
                    pvalueCutoff = 0.05,
                    pAdjustMethod = "BH",
                    qvalueCutoff = 0.2,
                    readable = TRUE)
    ego_results_filt[[stage]] <- ego
    if (!is.null(ego) && nrow(ego@result) > 0) {
      message(paste("Nombre de GO terms enrichis pour", stage, ":", nrow(ego@result)))
    } else {
      message(paste("Aucun GO term significatif trouvé pour", stage))
    }
    ego_simpl_filt[[stage]] <- simplify(ego)
  }
  
  # Traitement des feuilles "_spe" (uniquement si elles existent)
  if (length(sheets_spe) > 0) {
    ego_results_spe <- list()
    ego_simpl_spe <- list()
    for(stage in sheets_spe) {
      DEG_data <- read.xlsx(deg_path, sheet = stage, rowNames = FALSE)
      if (nrow(DEG_data) <= 1) {
        message(paste("Feuille", stage, "est vide. Ignorée."))
        next
      }
      DEG_data$gene <- as.character(DEG_data$gene)
      deg_map <- DEG_data %>% left_join(gff_ID, by = c("gene" = "geneName"))
      
      deg_entrez <- na.omit(deg_map$GeneID)
      
      ego <- enrichGO(gene = deg_entrez,
                      universe = universe_entrez,
                      OrgDb = org.ast,
                      keyType = "ENTREZID",
                      ont = "BP",
                      pvalueCutoff = 0.05,  
                      pAdjustMethod = "BH",
                      qvalueCutoff = 0.2,
                      readable = TRUE)
      ego_results_spe[[stage]] <- ego
      if (!is.null(ego) && nrow(ego@result) > 0) {
        message(paste("Nombre de GO terms enrichis pour", stage, ":", nrow(ego@result)))
      } else {
        message(paste("Aucun GO term significatif trouvé pour", stage))
      }
      ego_simpl_spe[[stage]] <- simplify(ego)
    }
  } else {
    message("Aucune feuille '_spe' trouvées.")
  }
  
  result <- list(
    ego_results_all = ego_results_all,
    ego_results_filt = ego_results_filt,
    ego_simpl_all = ego_simpl_all,
    ego_simpl_filt = ego_simpl_filt
  )
  
  # Ajouter les listes _spe uniquement si elles existent
  if (exists("ego_results_spe")) {
    result$ego_results_spe <- ego_results_spe
    result$ego_simpl_spe <- ego_simpl_spe
  }
  
  return(result)
}

save_results_to_excel <- function(results_list, file_name) {
  wb <- createWorkbook()
  for (stage in names(results_list)) {
    result <- results_list[[stage]]
    
    if (is.null(result) || nrow(result@result) == 0) {
      next
    }
    df <- result@result

    addWorksheet(wb, sheetName = stage)
    writeData(wb, sheet = stage, x = df, startRow = 1, startCol = 1)
  }

  saveWorkbook(wb, file_name, overwrite = TRUE)
}




save_all_results_to_excel <- function(enrichment_results, output_dir, prefix) {
  if (!dir.exists(output_dir)) {
    dir.create(output_dir, recursive = TRUE)
  }
  
  subsets <- list(
    list(name = "all", data = enrichment_results$ego_results_all),
    list(name = "filt", data = enrichment_results$ego_results_filt),
    list(name = "spe", data = enrichment_results$ego_results_spe),
    list(name = "simplify_all", data = enrichment_results$ego_simpl_all),
    list(name = "simplify_filt", data = enrichment_results$ego_simpl_filt),
    list(name = "simplify_spe", data = enrichment_results$ego_simpl_spe)
  )
  
  for (subset in subsets) {
    if (length(subset$data) > 0) {
      file_name <- file.path(output_dir, paste0("GO_analysis_", prefix, "_", subset$name, ".xlsx"))
      save_results_to_excel(subset$data, file_name)
    } else {
      message(paste("La liste", subset$name, "est vide. Aucun fichier Excel créé."))
    }
  }
}

save_all_dotplots <- function(enrichment_results, output_pdf) {
  pdf(output_pdf, width = 12, height = 8)
  
  # Définir l'ordre des sous-ensembles
  subset_order <- list(
    list(name = "all", data_key = "ego_results_all"),
    list(name = "filt", data_key = "ego_results_filt"),
    list(name = "spe", data_key = "ego_results_spe"),
    list(name = "simplify_all", data_key = "ego_simpl_all"),
    list(name = "simplify_filt", data_key = "ego_simpl_filt"),
    list(name = "simplify_spe", data_key = "ego_simpl_spe")
  )
  
  for (subset_info in subset_order) {
    name <- subset_info$name
    data_key <- subset_info$data_key
    
    if (!is.null(enrichment_results[[data_key]]) && length(enrichment_results[[data_key]]) > 0) {
      merged <- merge_result(enrichment_results[[data_key]])
      p <- dotplot(merged, color = "p.adjust", title = paste("Dotplot:", name), label_format = 45)
      print(p)
    }
  }
  
  dev.off()
}

```


```{r }

datasets <- list(
  list(name = "name_of_your_tissue1", path = ".../DEG_tissue1_by_stage.xlsx", output_dir = "results_path/"),
  list(name = "name_of_your_tissue2", path = ".../DEG_tissue2_by_stage.xlsx", output_dir = "results_path/")
)

for (dataset in datasets) {
  print(dataset$name)
  enrichment_results <- enrichment(dataset$path)

  save_all_results_to_excel(enrichment_results, dataset$output_dir, dataset$name)

  pdf_path <- file.path(dataset$output_dir, paste0("Dotplot_", dataset$name, "_results.pdf"))
  save_all_dotplots(enrichment_results, pdf_path)
}
```
